---
title: "Tidy Model"
author: "Ryan Bell"
date: "2025-02-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('tidyverse') ; library('modeldata')

```

## Introduction

This R Markdown shows my abilities to use models for data analysis using data set collected on crickets.


```{r load data, include = FALSE}
data(crickets, package = "modeldata")
names(crickets)
```

## What is that sound?

In this report, we examine what factors predict a cricket's chirp rate.


```{r summary, echo = FALSE}
summary(crickets)
ggplot(crickets, aes(x = rate)) +
  geom_histogram(bins = 8) + 
  ggtitle("Distribution of the chirp rate of crickets") +
  xlab('Chirp rate (per min.)')
```


## Temperature affects chirp rate

```{r temp, echo= FALSE}
ggplot(crickets, aes(x = temp, y = rate)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')

temp_lm <- lm(rate ~ temp, crickets)

summary.lm(temp_lm)
```

Based on a scatter plot of temperature and chirping and a correlation test, it seems that as temperature increases one degree, the rate of chirping also increases about 4.2 chirps per minute.

## Species-specific effects of temperature on chirping

```{r species, echo= FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate for two species of crickets") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')

species_lm <- lm(rate ~ temp + species, crickets)

summary.lm(species_lm)

```


The plot above shows a comparison of chirp rate between two species of crickets. Showing that the exclamationis species has a higher chirp rate compared to the niveus. However, the species in red has a higher chirp rate they were not tracked at a lower temperature compared to the niveus. This may be interesting to look more into as one cricket species could prefer certain temperatures and would remain inactive during others.

## Interactions

```{r species histogram, echo=FALSE}
ggplot(crickets, aes(x=rate, fill = species)) +
  geom_histogram(position = 'identity', alpha = 0.7, bins = 8) +
  ggtitle("Distribution of the chirp rate of crickets") +
  xlab('Chirp rate(per min.)')

```


```{r interactions, echo=FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate for two species of crickets") +
  ylab('Chirp rate(per min.)') +
  xlab('Temperature (Celsius)')


species_x_temp_lm <- lm(rate ~ temp + species + temp:species, crickets)
summary.lm(species_x_temp_lm)

anova(species_lm, species_x_temp_lm)
  
```

Checked for further interactions but decided to stay within the species model

## Predicting the worth of a Diamond

Without detailed instruction, try to build a linear model of at least three variables to predict the worth of a diamond.

```{r load data, include = FALSE}
data(diamonds, package = "ggplot2")
name(diamonds)


summary(diamonds)

```

Loaded diamond data

```{r temp, echo=False}
ggplot(diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  labs(title = "Effect of Carat & Cut on Diamond Price",
       x = "Carat",
       y = "Price")
```

Comparing each of the target variables (Carat, Color, and Cut) by Price shows significant to find differences between each of the cuts depending on their specific carat and cut.
